---
title: "R Notebook"
output: html_notebook
---

## Looking at DATA

```{r}
# plants is a dataset
class(plants)
```
```{r}
# Since it is a data.frame(), it is two dimensional

dim(plants)

nrow(plants)
ncol(plants)
```
```{r}
# how much space it occupies
object.size(plants)
```
```{r}
# variable names
names(plants)
```
```{r}
# take a peek at the data (default is 6 rows)
head(plants)
```
```{r}
# list 10 rows
head(plants, 10)
```
```{r}
# see last 15 rows
tail(plants, 15)
```
```
NAs are the place holders for the missing values.
```
```{r}
# see how the data is distributed and missing values
summary(plants)

# summary shows the basic stats for numeric variables but not for the character vectors
```
```{r}
# use table to see the distribution for the character vectors
table(plants$Active_Growth_Period)
```
```{r}
# anyway best concise option to see the structure of the data or any R object and function
str(plants)
```

# simulation
```{r}
# sample() generates random numbers

# simulate rolling four six-sided dice 
# with replacement: same number can show up more than once
sample(1:6, 4, replace=TRUE)
```

```{r}
# sample 10 numbers between 1 and 20 with replacement 
sample(1:20, 10)

# if size argument is not specified, then R assumes the same size of the source vector
```
```{r}
sample(LETTERS)
```

Let the value 0 represent tails and the value 1 represent heads.
Use sample() to draw a sample of size 100 from the vector c(0,1), with replacement. Since the coin is unfair, we must attach specific probabilities to the values 0 (tails) and 1 (heads) with a fourth argument, prob = c(0.3, 0.7). Assign the result to a new variable called flips.
```{r}
flips <- sample(c(0,1), 100, replace=TRUE, prob = c(0.3,0.7))
```
```{r}
# count the number of 1s
sum(flips)
```
```{r}
?rbinom
```
A binomial random variable represents the number of 'successes' (heads) in a given number of independent 'trials' (coin flips). Therefore, we can generate a single random variable that represents the number of heads in 100 flips of our unfair coin using rbinom(1, size = 100, prob = 0.7). Note that you only specify the probability of 'success' (heads) and NOT the probability of 'failure' (tails). Try it now.
```{r}
rbinom(1, size = 100, prob = 0.7)
```


```{r}
# if we want to see all of the 0s and 1s, we can request 100 observations, each of size 1, with success probability of 0.7.
flips2 <- rbinom(100, size = 1, prob = 0.7)

sum(flips2)
```
#  Normal Distribution
```{r}
# generate 10 random numbers from standart normal 
rnorm(10)
```

```{r}
rnorm(10, 100, 25)
```

# poisson
```{r}
#Generate 5 random values from a Poisson distribution with mean 10.

rpois(5, 10)
```
```{r}
# what if we want to simulate 100 *groups* of random numbers, each containing 5 values generated from a Poisson distribution with mean 10?

my_pois <- replicate(100, rpois(5, 10))
```
replicate() created a matrix, each column of which contains 5 random numbers generated from a Poisson distribution with mean 10.

```{r}
# find the mean of each column
cm <- colMeans(my_pois)
```
```{r}
# see central limit theorem at work
hist(cm)
```
## graphics
```{r}
data(cars)

# documentation for cars
?cars
```
```{r}
plot(x=cars$speed, y=cars$dist)
```
```{r}
plot(x=cars$speed, y=cars$dist, xlab = "Speed")
```
```{r}
plot(x=cars$speed, y=cars$dist, xlab = "Speed", ylab = "Stopping Distance")
```

```{r}
plot(x=cars$speed, y=cars$dist, main="My Plot")
```

```{r}
# subtitle
plot(cars, sub="My Plot Subtitle")
```
```{r}
#color red
plot(cars, col=2)
```
```{r}
# Plot cars while limiting the x-axis to 10 through 15
plot(cars, xlim = c(10, 15))
```
```{r}
# change the shape of the symbols
?points

plot(cars, pch = 2)
```

#load mtcars
```{r}
data(mtcars)
```
```{r}
?boxplot
```
```{r}
boxplot(formula = mpg ~ cyl, data = mtcars)
```











