---
title: "R Notebook"
output: html_notebook
---

## MOre on ggplot2

* data frame
* aesthetic mappings; color, size
* geoms: points, lines, shapes
* facets
* stats
* scales
* coordinate system


* Plots are built up in layers
    - Plot the data
    - overlay a summary
    - Metadata and annotation
    
```{r}
library(readxl)
library(dplyr)
library(tidyverse)
library(ggplot2)
```


```{r}
edu <- read_xlsx(path = "./data/education.xlsx")
df <- 
    read_xlsx(path = "./data/output.xlsx") %>%
    full_join(edu, by = c("con3code", "year")) %>%
    mutate(shangai = as.factor(shangai)) %>%
    drop_na()
names(df)
```
### Basic Plot

```{r}
qplot(x = enrollrate8, y = attainrate8,
      data = df,
          facets = . ~ factor(shangai), 
          geom = c("point", "smooth"),
          method = "lm")
```

```{r}
g <- ggplot(df, aes(x = enrollrate8, y = attainrate8))

summary(g)
```
```{r}
print(g)
```

##### You can add objects

```{r}
g + geom_point()
```

### loess smooth

```{r}
g + geom_point() + geom_smooth()
```
#### linear model
```{r}
g + geom_point() + geom_smooth(method = "lm")
```
#### Facet is on factor variables

labeling is important for factor variables

```{r}
g + geom_point()  + geom_smooth(method = "lm") +
    facet_grid(. ~ shangai)
```

## Annotation

labels: xlab, ylab, ggtitle
theme(legend.position = "none")

theme_gray()
theme_bw()

```{r}
g + geom_point(color = "steelblue",
               size = 4,
               alpha = 1/2)
```

```{r}
g + geom_point(aes(color = shangai),
               size = 4,
               alpha = 0.5)
```

### modify labels

```{r}
g + geom_point(aes(color = shangai),
               size = 2,
               alpha = 1) +
    labs(title ="Countries") +
    labs(x = "Enrollment rate",
         y = "Attainment Rate")
```

#### Modify smooth function

```{r}
g + geom_point(aes(color = shangai),
               size = 2,
               alpha = 1) +
    labs(title ="Countries") +
    labs(x = "Enrollment rate",
         y = "Attainment Rate") +
    geom_smooth(size = 4, 
                linetype = 3,
                method = "lm",
                se = F)
```



#### Modify theme

```{r}
g + geom_point(aes(color = shangai),
               size = 2,
               alpha = 1) +
    theme_bw(base_family = "Times")
```

#### Note on Axis and comparison

same and graph approach, but outlier is not shown in base plot

```{r}
test <- tibble(x = 1:100, y = rnorm(100))
# outlier created
test[50, 2] <- 100

plot(test$x, test$y, type = "l", ylim = c(-3, 3))
```


```{r}
g <- ggplot(test, aes(x = x, y = y))
g + geom_line()
```

#### Now modify y limit

Here outlier is not missing

```{r}
g + geom_line() + coord_cartesian(ylim = c(-3, 3))
```

#### Outlier is missing

```{r}
g + geom_line() + ylim(-3, 3)
```

### Complex example

#### prepare data
create factor variable

```{r}
df2 <- filter(df, year == 2016)
cutpoints <- quantile(df$patent_app_res, seq(0, 1, length = 5), 
                      na.rm = T)

df$new_uni <- cut(df$patent_app_res, cutpoints)

levels(df$new_uni)
```

```{r}
g <- ggplot(df, aes(x = enrollrate8,
                    y = attainrate8))

g + geom_point(alpha = 0.3) +
    facet_wrap(shangai ~ new_uni, nrow = 3, ncol = 3) +
    geom_smooth(method = "lm", se = FALSE, col = "steelblue") +
    theme_bw(base_family = "Avenir", base_size = 10) +
    labs(x = "Enrollment Rate - 8") +
    labs(y = "Attainment Rate - 8") +
    labs(title = "Last graph")
    
```






