---
title: "R Notebook"
output: html_notebook
---

```{r}
rm(list=ls())
library(ggplot2)
library(dplyr)
```

## Project - 2

Exploring the Data

```{r}
nei <- readRDS("data/summarySCC_PM25.rds")
scc <- readRDS("data/Source_Classification_Code.rds")
```
PM2.5 Emissions Data 
(summarySCC_PM25.rds): This file contains a data frame with all of the PM2.5 emissions data for 1999, 2002, 2005, and 2008. For each year, the table contains number of tons of PM2.5 emitted from a specific type of source for the entire year. Here are the first few rows.

* fips: A five-digit number (represented as a string) indicating the U.S. county
* SCC: The name of the source as indicated by a digit string (see source code classification table)
* Pollutant: A string indicating the pollutant
* Emissions: Amount of PM2.5 emitted, in tons
* type: The type of source (point, non-point, on-road, or non-road)
* year: The year of emissions recorded

### Question 1

Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.

```{r}
str(nei)
names(scc)
```

```{r}
unique(nei$Pollutant)
table(nei$year)
```
```{r}
summary(nei$Emissions)

nei_total <- aggregate(Emissions ~ year,  data = nei, FUN = sum ) 
nei_total
```
```{r, warning=FALSE}
# Open a png file
#png(filename = "plot1.png",
#    width = 480,
#    height = 480)

nei_total <- aggregate(Emissions ~ year,  data = nei, FUN = sum )

with(nei_total,
     plot(y = log10(Emissions), x =year,
     xlab = "Years", ylab = "Log(10) of Emissions",
     main = "Total Emissions",
     type = "b", col = "blue",
     lty = 1, lwd =1))

# Close the png file
# dev.off() 
```


### Question 2
Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008? Use the base plotting system to make a plot answering this question.

```{r, warning=FALSE}
# take a subset of nei, Baltimore city
baltimore <- subset(nei, fips =="24510")
baltimore_total <- aggregate(Emissions ~ year,  data = baltimore, FUN = sum )

# Open a png file
#png(filename = "plot2.png",
#    width = 480,
#    height = 480)

with(nei_total,
     plot(y = log10(Emissions), x =year,
     xlab = "Years", ylab = "Log(10) of Emissions",
     main = "Total Emissions - Baltimore City",
     type = "b", col = "forestgreen",
     lty = 1, lwd =1))

# Close the png file
#dev.off() 
```
### Question 3

Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999–2008 for Baltimore City? Which have seen increases in emissions from 1999–2008? Use the ggplot2 plotting system to make a plot answer this question.

```{r}
tapply(baltimore$Emissions, baltimore$type, mean, na.rm=TRUE)
```


```{r}
library(ggplot2)
qplot(x = year, y = Emissions, data = baltimore, color = type)
```
```{r}
qplot(x = year, y = log(Emissions), data = baltimore, facets = .~type)
```

```{r}
qplot(x = year, y = log(Emissions), data = baltimore, facets = type ~ .) +
    geom_smooth(method = "lm")
```
```{r}
qplot(x = year, y = log(Emissions), data = baltimore, color = type) +
    geom_smooth(method = "lm")
```


```{r, warning=FALSE}
baltimore_total2 <- aggregate(Emissions ~ year + type,  data = baltimore, FUN = sum )
baltimore_total2
```


```{r, warning=FALSE}
# Open a png file
# png(filename = "plot3.png",
#    width = 480,
#    height = 480)

ggplot(baltimore_total2, aes(x = year, y = Emissions)) +
    geom_line(aes(color = type)) +
    facet_grid(vars(type)) +
    labs(title ="Total Emissions - Baltimore City")

# Close the png file
# dev.off() 
```

```{r, fig.height=2}
ggplot(baltimore,aes(year, log(Emissions))) + 
    geom_boxplot() + facet_grid(type ~ .)
```

### Question 4

Across the United States, how have emissions from coal combustion-related sources changed from 1999–2008?

```{r}
unique(scc$EI.Sector)
```

```{r}
coal <- grepl("[Cc]oal", scc$EI.Sector)
sum(coal)
```
```{r}
coal2 = grepl("[Cc]oal", scc$Short.Name) & grepl("[Cc]omb", scc$Short.Name)
sum(coal2)
```

```{r}
coal_scc <- scc$SCC[coal2]
nei_coal <- nei[nei$SCC %in% coal_scc, ]

nei_total4 <- aggregate(Emissions ~ year ,  data = nei_coal, FUN = sum )
nei_total4
```

```{r}
# Open a png file
# png(filename = "plot4.png",
#    width = 480,
#    height = 480)

ggplot(nei_total4, aes(year, Emissions)) + 
    geom_line()  +
    labs(title ="Total Emissions - Coal Combustion")

# Close the png file
# dev.off() 
```

### Questions 5

How have emissions from motor vehicle sources changed from 1999–2008 in Baltimore City?

```{r}
unique(scc$EI.Sector)
```

```{r}
motor = grepl("[vV]ehicle", scc$EI.Sector) 

scc[motor, ]
```

```{r}
vehicle_scc <- scc$SCC[motor]

# take a subset of nei, Baltimore city
baltimore <- subset(nei, fips =="24510")
baltimore_vehicle <- baltimore[baltimore$SCC %in% vehicle_scc, ]

baltimore_tot5 <- aggregate(Emissions ~ year ,  data = baltimore_vehicle, FUN = sum )

```

```{r}
# Open a png file
#png(filename = "plot5.png",
#    width = 480,
#    height = 480)

ggplot(baltimore_tot5, aes(year, Emissions)) + 
    geom_line()  +
    labs(title ="Total Emissions - Baltimore - Motor Vehicles")

# Close the png file
# dev.off() 
```

### Question 6

Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (\color{red}{\verb|fips == "06037"|}fips == "06037"). Which city has seen greater changes over time in motor vehicle emissions?

```{r}
nei_sample <-
    nei%>%
    filter(SCC %in% vehicle_scc) %>%
    filter(fips  %in% c("06037", "24510")) %>%
    mutate(year = year,
           city = recode_factor(fips, '06037' = "Los Angeles County",
                                '24510' = "Baltimore City")) %>%
    group_by(year, city) %>%
    summarise(Emissions = sum(Emissions),
              .groups = "drop")


ggplot(nei_sample, aes(year, log10(Emissions), color = city)) + 
    geom_line()  +
    labs(title ="Total Emissions - LA vs Baltimore City")

```

```{r}

# take a subset of nei, Baltimore city
LA <- subset(nei, fips =="06037")
LA_vehicle <- LA[LA$SCC %in% vehicle_scc, ]
```

```{r}
LA %>%
    select(year, Emissions) %>%
    group_by(year) %>%
    summarise(avg_emis = mean(Emissions, na.rm=T))
```
```{r}
baltimore %>%
    select(year, Emissions) %>%
    group_by(year) %>%
    summarise(avg_emis = mean(Emissions, na.rm=T))
```








