---
title: "R Notebook"
output: html_notebook
---

## Creating new Variable

```{r}
if(!file.exists("./data")){dir.create("./data")}

fileurl <- "https://data.baltimorecity.gov/api/views/k5ry-ef3g/rows.csv?accessType=DOWNLOAD"

#download.file(fileurl, destfile = ".data/restaurants.csv", method = "curl")

restData <- read.csv(file = "./data/Restaurants.csv", )
```

# Sequences

```{r}
s1 <- seq(1, 10, by=2) ; s1

s2 <- seq(1, 10, length = 3); s2

x <- c(1, 3, 8, 25, 100); seq(along = x)
```

# subsetting

```{r}
restData$nearMe = restData$nghbrhd  %in% c("Roland Park", "Homeland")

table(restData$nearMe)
```

## Binary Variables
```{r}
# ifelse(condition, value when true, value when false)

restData$zipWrong = ifelse(restData$zipcode < 21205, TRUE, FALSE )

table(restData$zipWrong, restData$zipcode < 21205)

```

## Categorical Variables

```{r}
restData$zipcode = as.numeric(restData$zipcode)
restData$zipGroups  = cut(restData$zipcode, breaks = quantile(restData$zipcode , na.rm = T))
table(restData$zipGroups)
```

```{r}
table(restData$zipGroups, restData$zipcode)
```

## Easier Cutting
```{r}
library(Hmisc)

restData$zipGroups  = cut2(restData$zipcode, g = 4)

table(restData$zipGroups)
```

# creating factor variables

```{r}
restData$zcf  <- factor(restData$zipcode)

restData$zcf[1:10]

class(restData$zcf)
```

## Levels of factor

You can change the order of levels
```{r}
yesno <- sample(c("yes", "no"), size = 10, replace = TRUE)

yesnofac = factor(yesno, levels = c("yes", "no"))

relevel(yesnofac, ref = "yes")

```


```{r}
as.numeric(yesnofac)
```

# mutate

```{r}
library(Hmisc); library(plyr)

restData2 = mutate(restData, zipGroups = cut2(zipcode, g = 4))

table(restData$zipGroups)
```

### Some Useful Transformations

```{r}
x = 3.475

round(x, digits = 2) # 3.48

signif(x, digits = 2) # 3.5
```
```{r}
ceiling(x)

floor(x)
```


abs()                   absolute value
sqrt()                  square root
ceiling()               uste tamamlama
floor()                 alta tamamlama
round(x, digits=n)      yuvarlama
signif(x, digits=n)     yuvarlama ama farkli
cos(), sin()
log()
exp()





