---
title: "R Notebook"
output: html_notebook
---

Get/Set working directory
```
getwd()
setwd()
```
```{r}
getwd()
A <- getwd()
```
```{r}
setwd("~/Desktop/Coursera")
```


# list files or directories in working directory
```{r}
# list.dirs()
# dir()
str(list.files)
```

```{r}
# relative: move down one folder

list.files(path = "./data/", pattern = ".csv")
```

```{r}
#relative: move up one folder : coursera
dir(path = "../")
```

```{r}
dir(path = "../Rprogramming", pattern = ".html")
```
```{r}
# move up twice: desktop
dir(path = "../../", pattern = ".pdf")
```

# Check for and creating directories

file.create()
file.exists()
file.remove()
file.rename(from, to)
file.append(file1, file2)
file.copy(from, to)
file.symlink(from, to)
file.link(from, to)

```{r}
list.files(pattern = "direct")
```
```{r}
file.exists("directoryname")  # logical
```

```{r}
# create a directory if it does not exist
if (!file.exists("data")) {dir.create("data")}
```

# Downloading a file
download.file()
```{r}
str(download.file)
```

# right click on link to download and save the link of the file

method curl is advised for mac users if URL is Https. But it worked even without

```{r}
fileurl <- "https://opendata.arcgis.com/datasets/afdaf8cca48a4bcea9282a781e9190a6_0.csv"

download.file(url=fileurl, destfile = "./data/salary.csv", method = "curl")
```

```{r}
# record date if data is updated regularly
datedownloaded <- date()
```





















