---
title: "R Notebook"
output: html_notebook
---

```{r}
edu <- read_excel("../data/1.education.xlsx")

movies <- c("SPYDERMAN", "BATMAN", "VERTIGO", "CHINATOWN")
```
### lapply() function

**lapply()** function is useful for performing operations on list objects and returns a list object of same length of original set. 

**lappy()** returns a list of the similar length as input list object, each element of which is the result of applying FUN to the corresponding element of list. Lapply in R takes list, vector or data frame as input and gives output in list.

```{r}
args(lapply)
```

```{r}
movies_lower <- lapply(movies, tolower) 
str(movies_lower)
```
##### list to vector
```{r}
unlist(movies_lower)
```

### sapply() function

sapply() function takes list, vector or data frame as input and gives output in vector or matrix. It is useful for operations on list objects and returns a list object of same length of original set. Sapply function in R does the same job as lapply() function but returns a vector.

```{r}
movies_up <- sapply(movies, toupper) 
str(movies_up)
```
```{r}
df <- cars
str(df)
```


```{r}
sapply(df, min)
```
```{r}
sapply(df, function(x) {(min(x) + max(x))/2})
```

### tapply() function

tapply() computes a measure (mean, median, min, max, etc..) or a function for each factor variable in a vector. It is a very useful function that lets you create a subset of a vector and then apply some functions to each of the subset.

```{r}
str(tapply)
```

```{r}
df1 <- iris
str(df1)
```
```{r}
tapply(df1$Sepal.Length, df1$Species, mean)
```

edu <- 
edu data.frame(sapply(edu, as.numeric))

#3
train <- lapply(train, as.numeric)
train <- as.data.frame(do.call(cbind, train1))





