---
title: "R Notebook"
output: html_notebook
---

### Reading from TWITTER API

Be careful about what app in twitter you use, version 1.1 or version2.

there are packages to get data:
* retweet
* twitteR

httr package: GET, POST, PUT, DELETE
httr package: Facebook, Google, Twitter, Github

```{r, warning=FALSE}
library(httr)
library(jsonlite)
library(dplyr)

myBearerToken <- "AAAAAAAAAAAAAAAAAAAAAKRrQQEAAAAA%2FMPityrex8UayEwlWCszgTmrjJ0%3DIBxhgozO5Sey01KKP3s5azhiO4AWQoNbV8fDbiFb0dFxI0RLXa"
myApiKey <- "3A00h9XLepNUmphHZNPPEIrU2"
myApiSecretKey <- "9i5uif6OZJ2pI0l4s1dYECrBZQFPvArKRuoG42HhjiUML4aIqk"
myAccessToken <- "784096391526227968-hBXEZekqRYY5a8N3VrcfSsjXgAeoaEd"
myAccessTokenSecret <- "UOgGELiuBGUQh2kV1KxH0jWD5veUY84UTvfFl8oYJucwx"
```



The Bearer Token is what allows you to authenticate to the Twitter API and start making requests.
You can use the package httr to make HTTP requests to the Twitter API.
You will also need to install jsonlite to work with our JSON object and dplyr for data manipulation.

```{r}
Sys.setenv(BEARER_TOKEN = myBearerToken)
```

Grab the Bearer Token you pulled from your App, and pass that into your headers for authentication. 
```{r}
bearer_token <- Sys.getenv(myBearerToken)
headers <- c(`Authorization` = sprintf('Bearer %s', bearer_token))
```

Once you got your authentication set up, define the parameters of your request. By default, you will get back the id, name, and username of each user you get back. You can make adjustments to this payload by adding additional fields and expansions. For this example, you will want the profile bio of the user which is requested using user.fields=description, and an expansion that contains the pinned Tweet of the user.
```{r}
params <- list(`user.fields` = 'description',
               `expansion` = 'pinned_tweet_id')
```


Now you are ready to format your URL with the Twitter handle, also known as account, you are looking to get more information about. Use the readline method to allow this sample to be reusable. After you type the handle you want to look at, format your URL to contain the handle you define by replacing $USERNAME with the desired Twitter handle. 
```{r}
handle <- readline('DavutLava')

url_handle <-
  sprintf('https://api.twitter.com/2/users/by?usernames=%s', handle)
```

At this point, use the httr package to make a GET request to the URL you just 
created, pass in our authentication credential via the header, and pass in the
parameters you defined. You can save the response as a text object in the 
variable obj and print this out to view the result of the request you made.

```{r}
response <-
  httr::GET(url = url_handle,
            httr::add_headers(.headers = headers),
            query = params)
obj <- httr::content(response, as = "text")
print(obj)
```

# Working with JSON
One of my favorite ways to work with a JSON is to use a data frame, which allows you to easily access complex nested data. To do this, use the fromJSON method of the jsonlite package to flatten your file to allow the fields to be in the same object. Then, pass that object into a data frame. Now you are ready to view this data frame.

```{r}
json_data <- fromJSON(obj, flatten = TRUE) %>% as.data.frame
View(json_data)
```

You can access the fields of the data from the data frame and pass them into a string that has the handle, the username, and bio.

```{r}
final <-
  sprintf(
    "Handle: %s\nBio: %s\nPinned Tweet: %s",
    json_data$data.username,
    json_data$data.description,
    json_data$includes.tweets.text
  )
```

Use cat instead of print to view the object with newlines between each field you are pulling.

```{r}
cat(final)
```

# Some Other way to use Twitter API
```{r}
library(rtweet)
library(tidytext)

appname <- "LearningR"
myApiKey
myApiSecretKey
```

Authentication
```{r}
twitter_token <- create_token(
    app = appname,
    consumer_key = myApiKey,
    consumer_secret = myApiSecretKey,
    access_token = myAccessToken,
    access_secret = myAccessTokenSecret)
```

# post a tweet
```{r}
post_tweet("Look, Im tweeting using R @DavutLava")
```

```{r}
twitter_token
```
# Search Tweets
```{r}
rstats_tweets <- search_tweets(q = "#rstats",
                               n = 500)
```

# below is version 1.1 (Coursera)
```{r}
myapp = oauth_app("twitter", key = oldAPIkey,
                  secret = oldAPIsecretKey)
sig = sign_oauth1.0(myapp, token = oldAccessToken,
                    token_secret = oldAccessTokenSecret)
# specific URL: 1.1 is version; statuses is the theme of data;
# and data name; json is the only data type
sampleURL <- "https://api.twitter.com/2/tweets/search/stream"

homeTL = GET(sampleURL, sig)
## Converting the JSON object

```{r}
library(jsonlite)
json1 = content(homeTL)
json2 = jsonlite::fromJSON(toJSON(json1))
json2[]
```






