---
title: "Knitr Package"
author: "dea"
date: "8/21/2021"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Setting Global Options

This is expected to be at the top

```{r setoptions, echo=FALSE}
# opts_chunk$set(echo = FALSE, results = "hide")
```

### Notebook vs Rmarkdown

* They can create same outputs  
* Use same syntax  
* For output in Notebook, you need to run all chunks of codes to appear in the output  
* In Rmarkdown, when you knit, automatically it runs all the code and produce the output.


My First knitr Document
=============================================

This is some text

Here is a code chunk
```{r}
set.seed(1234)
x = rnorm(100)
mean(x)
```

### More Complicated Way

```
library(knitr)
setwd(getwd())
knit2html("document.Rmd")
browseURL("document.html")
```

# Another Example

## Intro

```{r firstchunk, echo=FALSE, results='hide'}
x = rnorm(100)
summary(x)
```
R objects can be used as macros in the text

```{r}
x = Sys.time()
y = mean(rnorm(100))

format(x, "%Y-%m-%d, %a")
```

System time is `r x`. My favorite number is `r y`.

### incorporate Graphics

```{r scatterplot, fig.height=4}
x = rnorm(100); y = x +rnorm(100, sd = 0.5)

par(mar = c(5, 4, 1, 1), las = 1)
plot(x, y, main = "My Simulated Data")
```

### Making Tables with xtable package  

```{r fitmodel}
data("airquality")

fit = lm(Ozone ~ Wind + Temp + Solar.R, data = airquality)

summary(fit)
```

Making a table of regression coefficients

```{r showtable, results='asis'}
library(xtable)
xt = xtable(summary(fit))

```

For pdf output

```{r showtable1}
print(xt, type = "latex")
```

Html format
```{r showtable2}
print(xt, type = "html")
```



`results`: "asis", "hide"
`echo`: TRUE / FALSE
`fig.height`: numeric
`fig.width`: numeric

**important tip**  
Some code chunks take forever to run. You do not have to run it every time.

`cache=TRUE`







