---
title: "R Notebook"
output: html_notebook

knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html_output") })
---

### Reproducible Research DOs and DONTs
1. DOs
garbage in garbage out
Coherent, focused question
good collaborators
interesting topic
keep software, package track of versions

### Use version control system
* software like Github, Bitbucket, SourceForge to publish results
* it will slow things down
* add small changes each time
* track snapshots or revert to old
*


2. DON'Ts
dont do hand editing
downloading data from web by clicking the links
dont use point and click software


```{r}
url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/00275/Bike-Sharing-Dataseet.zip"

download.file(url, destfile = "data.zip")
```

### Advantages
* you have the specific url
* downloaeded to local machine
* file name
* location is specific

### Important: Dont save output alone

* It is important that keep output only with code. Otherwise, you will have outputs that may not be reproducible

* Every intermediate file should be documented

### DO: Think about the entire pipeline

* Data analysis is lengthy process; it is not just tables, figures, reports

* Raw data ---> processed data ---> analysis ---> report

* How you got to the end is as important as the end itself

### The `cacher` package for R


### Simple Analysis

```{r}
library(datasets)
library(stats)
```

```{r}
data("airquality")
```
```{r}
fit <- lm(Ozone ~ Wind + Temp + Solar.R, 
          data = airquality)
summary(fit)
```

```{r, fig.height=4, fig.width=4}
par(mfrow = c(2, 2))
plot(fit)
```


