---
title: "R Notebook"
output: html_notebook
---

## Debugging Tools - Diagnosing the Problem

message: notification message, execution continue

warning: not fatal but something wrong

error: fatal and execution stops

condition: something unexpected can occur
```{r}
# warning
log(-1)
```
```{r}
printmessage <- function(x) {
    if(x>0) print("x is greater than zero")
    else print("x is less than or equal to zero")
    invisible(x)
}

printmessage(1)
```
```{r}
printmessage(NA)
```
```{r}
printmessage2 <- function(x) {
    if(is.na(x)) print("x is missing value")
    else
    if(x>0) print("x is greater than zero")
    else print("x is less than or equal to zero")
    invisible(x)
}

x <- log(-1)
```
```{r}
printmessage2(x)
```
# something is wrong
user mistakes

what are your expectations, can you reproduce the problem

## Debugging tools

traceback : prints out function call stack
debug : flags a  function for debug mode
browser: 
trace
recover
print or cat statements
```{r}
mean(y)
```
```{r}
traceback()
```
```{r}
debug(lm)
lm(y~x)

# press n to see the next argument
```

```{r}
options(error = recover)

read.csv("nosuchfile")
```




